Using Visual Analytics to Explain\\[1ex] Black-Box Machine Learning

### 2018-03-25 tex/introduction/machine_learning.tex
restructuring

\begin{figure}
\centering
\includegraphics[height=8em,valign=t]{tex/introduction/age.png}
\caption{
An influence function for ``age" of a Generalized Additive Model \cite{Caruana:2015:IMH:2783258.2788613}.
}
\label{figs:age}
\end{figure}

@@

\section{Visual Analytics}
Gaining meaningful insights about large data sets is hard.
While computing various statistics about the data is often helpful it might oversimplify or in some cases even mislead information about the data.
One example data set often used to demonstrate this is Ancombe's quartet \todo{cite ancombe}.
The quartet consists of four different data sets that all have the same mean and sample variance for both axes, the same correlation between the axes, and the same regression lines.
However, plotting the values of the data sets reveals that each one of those data sets obeys a different, unique characteristic (see Figure~\ref{fig:ancombe}).

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth,valign=t]{tex/introduction/ancombe.png}
\caption{
Plots of the four data sets of Ancombe's quartet \todo{cite wikipedia}. All data sets have the same mean and sample variance for both axes, the same correlation between the axes, and the same regression lines.
}
\label{figs:ancombe}
\end{figure}

>>>>

\begin{figure}
\centering
\includegraphics[height=8em,valign=t]{tex/introduction/age.png}
\caption[Influence function for ``age" from a Generalized Additive Model.]{
An influence function for ``age" from a Generalized Additive Model \cite{Caruana:2015:IMH:2783258.2788613}.
}
\label{figs:age}
\end{figure}

@@

\section{Visual Analytics}
Gaining meaningful insights about large data sets is hard.
While computing various statistics about the data is often helpful it might oversimplify or in some cases even mislead information about the data.
One example data set often used to demonstrate this is Ancombe's quartet \todo{cite ancombe}.
The quartet consists of four different data sets that all have the same mean and sample variance for both axes, the same correlation between the axes, and the same regression lines.
However, plotting the values of the data sets reveals that each one of those data sets obeys a different, unique characteristic (see Figure~\ref{figs:ancombe}).

\begin{figure*}[t]
\centering
\includegraphics[width=0.9\textwidth,valign=t]{tex/introduction/ancombe.png}
\caption[Ancombe's quartet.]{
Plots of the four data sets of Ancombe's quartet\footnotemark. All data sets have the same mean and sample variance for both axes, the same correlation between the axes, and the same regression lines.
}
\label{figs:ancombe}
\end{figure*}
\footnotetext{Image source: Wikipedia}

### 2018-03-25 tex/introduction/motivation.tex
finish up ML section
use case
restructuring

\section{Motivation}
\todo{Motivation}
Modern machine learning algorithms have reached a degree of potential that lets the resulting models model almost any dataset.
But even though there are statistical measures to guarantee that the model is learning correctly without over fitting on the training data, the model with the highest accuracy is not always the most usable model or even semantically correct.
The following three examples aim to showcase this:

\par \noindent (1)
\todo{pneumonia example}
% \cite{Caruana:2015:IMH:2783258.2788613}

\begin{figure}
\centering
\includegraphics[height=6em,valign=t]{tex/introduction/pneumonia.png}
\includegraphics[height=8em,valign=t]{tex/introduction/age.png}
\includegraphics[height=8em,valign=t]{tex/introduction/asthma.png}
\caption{
\todo{TODO}
}
\label{figs:pneumonia}
\end{figure}

\par \noindent (2)
\todo{radio example}
% \cite{Bird:2002:ERI:1251972.1252349}
% https://people.duke.edu/~ng46/topics/evolved-radio.pdf

\begin{figure}
\centering
\includegraphics[height=10em]{tex/introduction/emradio.png}
\caption{
\todo{TODO}
}
\label{figs:emradio}
\end{figure}

@@

In order for a technique to be effectively useful for a human it needs to have few restrictions \todo{}
* reasons for black box
* gradient to explanation

>>>>

\section{Motivation}
Modern machine learning algorithms have reached a degree of potential that lets the resulting models model almost any data set.
But even though there are statistical measures to guarantee that the model is learning correctly without over-fitting on the training data, the model with the highest accuracy is not always the most usable model or even semantically correct.
The following three examples aim to showcase this:

\par \noindent (1)
Caruana~\etal\cite{Caruana:2015:IMH:2783258.2788613} built a Generalized Additive Model with pairwise interaction ($GA^2M$) to predict mortality risk for hospitalized patients with pneumonia, a potentially life-threatening lung infection. This area of machine learning is called predictive modeling and uses machine learning to predict a certain outcome, in this case mortality risk, based on values of several input features, in this case laboratory measurements, comorbidities, or patient demographics, and is trained from and applied to many instances. It is not feasible to individually check each prediction of the model manually. However, the machine learning algorithm $GA^2M$ used for this study is an intelligible model, that sacrifices parts of its predictive quality in favor of being understandable by humans. This enabled the authors to inspect how features contributed to the predicted outcome. One of their findings was that the model associated having asthma, a chronic lung disease, with a lower mortality rate for pneumonia patients. However, this combination of diseases is known to have a significantly \emph{increased} mortality rate. The authors double checked their data and found that this phenomenon actually occurred in their dataset and thus appeared as correct prediction in their model. After further investigation it became clear that this bias in the dataset stemmed from patients, with those conditions, being treated with intensive care thus lowering their mortality risk below the average. Outside of a hospital the risk of those patients would have been significantly higher. The model, however, only learnt about hospitalized patients which led to it being confident in this false association.

\begin{figure}
\centering
\includegraphics[height=6em,valign=t]{tex/introduction/pneumonia.png}
\includegraphics[height=8em,valign=t]{tex/introduction/age.png}
\includegraphics[height=8em,valign=t]{tex/introduction/asthma.png}
\caption[Influence functions of Pneumonia model.]{
Pneumonia is an infectious disease of the lungs that inflames air sacs and likely fills them with fluid, making it hard to breathe and potentially life-threatening (left\footnotemark).
The influence functions of the features ``age" (middle) and ``asthma" (right) of the model created by Caruana~\etal\cite{Caruana:2015:IMH:2783258.2788613}.
}
\label{figs:pneumonia}
\end{figure}
\footnotetext{Image source: Google}

\par \noindent (2)
Layzell and Bird \cite{Bird:2002:ERI:1251972.1252349} used machine learning to automatically design an oscillator, an electronic circuit that produces a periodic oscillating signal, using a physical programmable circuit rather than a simulated circuit. After the model completed the task, the scientists looked at the resulting electronic circuit and were surprised to find that instead of the conventional design, using a feedback loop of an amplifier and a filter, the model designed a makeshift radio that responded to the desired frequency and amplified radio signals emitted by nearby computers instead. The output signal of this electronic circuit had the correct frequency and amplitude according to the specification but did not create a valid result due to a latent bias in the available data.

\begin{figure}
\centering
\includegraphics[height=10em]{tex/introduction/emradio.png}
\caption{
The physical programmable circuit used by Bird and Layzell \cite{Bird:2002:ERI:1251972.1252349} for solving the task of automatically designing oscillators.
}
\label{figs:emradio}
\end{figure}
% https://people.duke.edu/~ng46/topics/evolved-radio.pdf

@@

% There has been research...specialized models
% In order for a technique to be effectively useful for a human it needs to have few restrictions \todo{}
* black-box vs. white-box
* reasons for black box
\todo{problem statement}

The above example illustrates the need for human understanding and supervision in machine learning and the incorrect correct result could easily be identified since the authors used a special, intelligible, model. However, not all models used in machine learning are, or even attempt to be, easily interpretable by human experts. To counteract this, visual analytics has been successfully used to visualize machine learning processes in order to understand and gain insights into predictive models. Visual analytics is the area of studying interactive graphical representations of abstract information with the help of computational and statistical procedures, in order to effectively analyze complex data. With regards to understanding the decision making of predictive machine learning models, there are two main strategies: white-box and black-box analysis \cite{class_signatures}.

% \section{White-box vs. Black-box Analysis}
In white-box analysis the main focus lies in communicating the internal structure of the model at hand. That is, the analysis is based on the assumption that knowing the full internal state of a model and being able to manually retrace decisions made by it, helps understanding the \emph{behavior} of the model in general.

In this work we will focus on black-box analysis whose focus is on communicating external behavior of a model. That is, no information about the model itself is utilized during the analysis but rather typically inferred from observing how the model reacts to carefully crafted inputs. Compared to white-box analysis there are several advantages.

As white-box analysis relies on the internal structure and thus the nature of a particular model type, such as the $GA^2M$ mentioned above, a method developed for one type might not be applicable to a different type. This limits the usefulness of white-box analysis as every new algorithm demands often an entirely new form of representation. Since black-box analysis is model independent findings are universally applicable.

\newcommand\Tstrut{\rule{0pt}{2.5ex}}
\begin{table}
     \begin{tabular}{cl|cl} 
     & \textbf{White-box} & & \textbf{Black-box} \\
     \hline
     \hline
    $\oplus$ & Deep understanding of decisions & $\ominus$ & External behavior only \Tstrut \\
    $\oplus$ & Always reflects the true & $\ominus$ & Not guaranteed to reflect model \Tstrut \\
    & decision making process & & decisions truthfully in all cases \\
    $\oplus$ & Full internal state available & $\ominus$ & Limited knowledge of \Tstrut \\
    & & & model specifics \\
    \hline
    $\ominus$ & Model specific techniques & $\oplus$ & Reusable techniques \Tstrut \\
    $\ominus$ & Complexity / interpretability & $\oplus$ & Complexity / interpretability \Tstrut \\ 
    & data and model dependent & & independent of data and model \\
    $\ominus$ & Switching to interpretable model & $\oplus$ & No performance considerations \Tstrut \\
    & incurs performance penalty & & necessary \\
    \end{tabular}
    \centering
    \vspace*{-0.5em}
    \caption[Comparison of black-box and white-box explanation strategies.]{Comparison of black-box and white-box explanation strategies. $\oplus$ indicates advantages of a technique while $\ominus$ indicates drawbacks.}
    \vspace*{-0.75em}
    \label{tab:blackvswhite}
\end{table}

Another drawback of white-box analysis is its scalability to the complexity of the model. Many solutions do not sufficiently scale as models become more complex. For example, a decision tree with 5 nodes, often used as example when explaining the technique, can easily be visualized and understood in a node-link representation. This representation fails to help understanding a decision tree with hundreds or thousands of nodes. For black-box analysis the internal complexity of a model is irrelevant.

On the other hand black-box analysis also has drawbacks. Especially the limited knowledge of the underlying model allows only for an approximate understanding of its behavior as testing the entire input space would be intractable.
